Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-01-30 manuscripts/TCS2022/tcs2022.tex
sorted by network size
finish conclusion

\begin{table}[!htb]
  \caption{Timing comparisons (in seconds) between \texttt{bioLQM} (\texttt{LQM}), \texttt{PyBoolNet} (\texttt{PBN}), \texttt{mpbn} and the four variants of \texttt{Trappist} on selected models from the literature.}
  % in bold if greater than or equal to 10
  \centering%
  \label{tab:result_real}
  \scalebox{0.8}{
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{\texttt{Trappist}}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & \texttt{mpbn} & SAT & CLP & ILP & ASP \\
    \midrule % n < 100
    \rownb & Arabidopsis\_thaliana~\cite{DesignPrinciplesGeneNetworks} & 15 & 8 & \textbf{0.10} & 0.06 & NM & 0.01 & - & - & 0.02 \\
    \rownb & Bcell~\cite{Dutta2019} & 73 & 72 & 0.23 & 0.04 & 0.08 & 0.06 & - & - & 0.05 \\
    \rownb & Cacace\_TdevModel~\cite{Cacace2020} & 61 & 28 & \textbf{1.29} & \textbf{5.67} & NM & 0.06 & - & - & 0.08 \\
    \rownb & Corral\_ThIL17diff~\cite{corral2021interplay} & 92 & $>1000$ & N/A & \textbf{107.57} & 0.76 & 0.56 & - & - & 0.16 \\
    \rownb & DNA\_damage~\cite{DesignPrinciplesGeneNetworks} & 26 & 16 & \textbf{0.24} & \textbf{0.33} & NM & 0.02 & - & - & 0.05 \\
    \rownb & Drosophila~\cite{RodrguezVega2014} & 52 & 128 & 0.33 & 0.05 & 0.07 & 0.06 & - & - & 0.05 \\
    \rownb & EMT~\cite{Rozum2021} & 69 & 268 & \textbf{39.22} & \textbf{1.01} & 0.20 & 0.12 & - & - & 0.05 \\
    \rownb & mast\_cell~\cite{aghamiri2020automated} & 73 & $>1000$ & N/A & 0.09 & 0.55 & 0.37 & - & - & 0.15 \\
    \rownb & FT-GRN~\cite{ChvezHernndez2022} & 23 & 32 & \textbf{DNF} & \textbf{DNF} & NM & 0.03 & - & - & 0.19 \\
    \rownb & hedgehog~\cite{DesignPrinciplesGeneNetworks} & 65 & $>1000$ & N/A & \textbf{DNF} & 0.50 & 0.34 & - & - & 0.33 \\
    \rownb & metastatic~\cite{DesignPrinciplesGeneNetworks} & 10 & 4 & \textbf{0.10} & 0.04 & NM & 0.01 & - & - & 0.02 \\
    \rownb & Pancreatic\_Cancer~\cite{DesignPrinciplesGeneNetworks} & 43 & $>1000$ & N/A & 0.11 & 0.36 & 0.17 & - & - & 0.06 \\
    \rownb & Pluripotent~\cite{DesignPrinciplesGeneNetworks} & 36 & 276 & 0.37 & 0.43 & NM & 0.07 & - & - & 0.06 \\
    \rownb & Rho-GTPases~\cite{DesignPrinciplesGeneNetworks} & 33 & 2 & 0.17 & 0.57 & \textbf{40.39} & 0.07 & - & - & 0.11 \\
    \rownb & Pluripotency~\cite{YachieKinoshita2018} & 36 & 440 & \textbf{DNF} & \textbf{DNF} & NM & 0.16 & - & - & 0.28 \\
    \rownb & p53\_high\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.38 & \textbf{1.76} & NM & 0.08 & - & - & 0.14 \\
    \rownb & p53\_low\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.41 & \textbf{1.76} & NM & 0.07 & - & - & 0.14 \\

    \midrule % 100 <= n < 200
    \rownb & angiofull~\cite{Weinstein2017} & 142 & $>1000$ & N/A & 0.17 & 1.06 & 0.88 & - & - & 0.23 \\
    %\rownb & COVID\_19 & 111 & - & N/A & - & - & - & - & - & - \\
    \rownb & EMT\_Mech~\cite{Sullivan2022} & 136 & 82 & \textbf{DNF} & \textbf{14.01} & 0.27 & 0.20 & - & - & 0.25 \\
    \rownb & EMT\_Mech\_TGFbeta~\cite{Sullivan2022} & 150 & 492 & \textbf{DNF} & \textbf{11.28} & 0.78 & 0.69 & - & - & 0.35 \\
    \rownb & MAPK~\cite{aghamiri2020automated} & 181 & $>1000$ & N/A & \textbf{13.58} & 1.76 & 1.51 & - & - & 0.27 \\
    \rownb & macrophage~\cite{DesignPrinciplesGeneNetworks} & 136 & $>1000$ & N/A & 0.54 & 1.09 & 0.84 & - & - & 0.27 \\
    \rownb & RA\_apoptosis~\cite{aghamiri2020automated} & 180 & $>1000$ & N/A & \textbf{DNF} & 1.43 & 1.55 & - & - & 0.19 \\
    \rownb & Adhesion\_CIP~\cite{guberman2020boolean} & 121 & 78 & \textbf{56.81} & \textbf{4.25} & 0.23 & 0.17 & - & - & 0.19 \\
    \rownb & angiogenesis~\cite{DesignPrinciplesGeneNetworks} & 141 & $>1000$ & N/A & 0.16 & 1.07 & 1.06 & - & - & 0.16 \\

    \midrule % 200 <= n
    \rownb & T-cell-co-receptor~\cite{DesignPrinciplesGeneNetworks} & 206 & $>1000$ & N/A & \textbf{DNF} & 1.52 & 2.26 & - & - & 0.35 \\
    \rownb & Snf1-pathway~\cite{Lubitz2015} & 202 & $>1000$ & N/A & 1.13 & 1.47 & 1.43 & - & - & 0.31 \\
    \rownb & Mycobacterium~\cite{DesignPrinciplesGeneNetworks} & 317 & $>1000$ & N/A & 0.42 & 2.36 & \textbf{4.91} & - & - & 0.44 \\
    \rownb & Leishmania~\cite{DesignPrinciplesGeneNetworks} & 342 & $>1000$ & N/A & \textbf{DNF} & 2.56 & \textbf{5.62} & - & - & 0.46 \\
    \rownb & TcellCheckPoint~\cite{hernandez2020computational} & 218 & $>1000$ & N/A & \textbf{4.99} & NM & 1.96 & - & - & 0.28 \\
    \rownb & Cholocystokinin~\cite{aghamiri2020automated} & 383 & $>1000$ & N/A & 0.36 & 2.99 & \textbf{4.81} & - & - & 0.37 \\
    \rownb & Alzheimer~\cite{aghamiri2020automated} & 762 & $>1000$ & N/A & \textbf{DNF} & NM & \textbf{18.21} & - & - & 0.79 \\

@@

The last observation is that for very large models (i.e., more than two thousand nodes) the proposed method did not manage to finish the Petri net conversion before the timeout, as marked by ``???'' in Table~\ref{tab:result_real}.
This points to the fact that our current choice of using a BDD-based translation to obtain that Petri net encoding, though it provides a small/efficient ASP might be too costly to handle the largest models.
In such a case, a more \emph{naive} encoding might provide a much larger ASP program, with many redundant rules, but easier/faster to obtain.
The evaluation of the feasibility of such strategy, and of its impact on smaller instances, remains to be done and is out of the scope of this first article.
Recognizing that a model is locally-monotonic and applying in that specific case dedicated strategies as those of \texttt{mpbn} might also be a partial solution as shown in the SN-5 case, but not for the turei-2016 model.

@@

%Though we benchmarked this new approach against state-of-the-art tools, there are many more evaluations that we plan to do in the future.
%First, the \texttt{bioLQM} platform that we are using is providing another implicant-based method using BDDs in \url{http://colomoto.org/biolqm/doc/tools-trapspace.html} and though we expect it to behave mostly like \texttt{PyBoolNet}, that remains to be checked.
%Note that this also raises the question of replacing altogether the \texttt{bioLQM} preprocessing step we use to obtain the Petri net encoding, since its use of BDDs might not be optimal for that step.
%The trade-off between a small Petri net, and hence small ASP, and the time it takes to compute it (in other words, the trade-off of allowing redundant constraints) has to be evaluated in depth.

% TODO: Compare among various possible methods and obtained insights
Second, the experimental results shown in this paper mostly expose the differences caused by the prime-implicant phase of \texttt{PyBoolNet}. There is much more information required to distinctly study the practical differences between our ASP encoding and that of \texttt{PyBoolNet}, and notably their size/efficiency ratio.
Hence, we plan to conduct experiments on more real-world models or maybe random models that can be randomly generated by using BoolNet~\cite{mussel2010boolnet}.
Such experiments should include a time-course of the number of ASP solutions found for proper comparison of the ASP encodings.

% TODO: can borrow many advents in the field of Petri nets
In addition, there are possibly other methods for computing maximal conflict-free siphons in Petri nets, like SAT/MaxSAT approaches~\cite{nabli2016enumerating}. Although these approaches do not directly support the maximal conflict-free siphon computation now, we plan to investigate them in the future. They could replace our ASP program if they outperform it.
However, the current method appears to already perform very well even on the biggest models we have considered.

>>>>

\begin{table}[!htb]
  \caption{Timing comparisons (in seconds) between \texttt{bioLQM} (\texttt{LQM}), \texttt{PyBoolNet} (\texttt{PBN}), \texttt{mpbn} and the four variants of \texttt{Trappist} on selected models from the literature.}
  % in bold if greater than or equal to 10
  \centering%
  \label{tab:result_real}
  \scalebox{0.8}{
  \begin{tabular}{rlrrrrrrrrr}
    \toprule
    & & & & & & & \multicolumn{4}{c}{\texttt{Trappist}}\\
    \cmidrule(rr){8-11}
    & model & \(n\) & \(|M|\) & LQM & PBN & \texttt{mpbn} & SAT & CLP & ILP & ASP \\
    \midrule % n < 100
    \rownb & metastatic~\cite{DesignPrinciplesGeneNetworks} & 10 & 4 & \textbf{0.10} & 0.04 & NM & 0.01 & - & - & 0.02 \\
    \rownb & Arabidopsis\_thaliana~\cite{DesignPrinciplesGeneNetworks} & 15 & 8 & \textbf{0.10} & 0.06 & NM & 0.01 & - & - & 0.02 \\
    \rownb & p53\_high\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.38 & \textbf{1.76} & NM & 0.08 & - & - & 0.14 \\
    \rownb & p53\_low\_dna~\cite{DesignPrinciplesGeneNetworks} & 16 & 1 & 0.41 & \textbf{1.76} & NM & 0.07 & - & - & 0.14 \\
    \rownb & FT-GRN~\cite{ChvezHernndez2022} & 23 & 32 & \textbf{DNF} & \textbf{DNF} & NM & 0.03 & - & - & 0.19 \\
    \rownb & DNA\_damage~\cite{DesignPrinciplesGeneNetworks} & 26 & 16 & \textbf{0.24} & \textbf{0.33} & NM & 0.02 & - & - & 0.05 \\
    \rownb & Rho-GTPases~\cite{DesignPrinciplesGeneNetworks} & 33 & 2 & 0.17 & 0.57 & \textbf{40.39} & 0.07 & - & - & 0.11 \\
    \rownb & Pluripotency~\cite{YachieKinoshita2018} & 36 & 440 & \textbf{DNF} & \textbf{DNF} & NM & 0.16 & - & - & 0.28 \\
    \rownb & Pluripotent~\cite{DesignPrinciplesGeneNetworks} & 36 & 276 & 0.37 & 0.43 & NM & 0.07 & - & - & 0.06 \\
    \rownb & Pancreatic\_Cancer~\cite{DesignPrinciplesGeneNetworks} & 43 & $>1000$ & N/A & 0.11 & 0.36 & 0.17 & - & - & 0.06 \\
    \rownb & Drosophila~\cite{RodrguezVega2014} & 52 & 128 & 0.33 & 0.05 & 0.07 & 0.06 & - & - & 0.05 \\
    \rownb & Cacace\_TdevModel~\cite{Cacace2020} & 61 & 28 & \textbf{1.29} & \textbf{5.67} & NM & 0.06 & - & - & 0.08 \\
    \rownb & hedgehog~\cite{DesignPrinciplesGeneNetworks} & 65 & $>1000$ & N/A & \textbf{DNF} & 0.50 & 0.34 & - & - & 0.33 \\
    \rownb & EMT~\cite{Rozum2021} & 69 & 268 & \textbf{39.22} & \textbf{1.01} & 0.20 & 0.12 & - & - & 0.05 \\
    \rownb & Bcell~\cite{Dutta2019} & 73 & 72 & 0.23 & 0.04 & 0.08 & 0.06 & - & - & 0.05 \\
    \rownb & mast\_cell~\cite{aghamiri2020automated} & 73 & $>1000$ & N/A & 0.09 & 0.55 & 0.37 & - & - & 0.15 \\
    \rownb & Corral\_ThIL17diff~\cite{corral2021interplay} & 92 & $>1000$ & N/A & \textbf{107.57} & 0.76 & 0.56 & - & - & 0.16 \\

    \midrule % 100 <= n < 200
     \rownb & Adhesion\_CIP~\cite{guberman2020boolean} & 121 & 78 & \textbf{56.81} & \textbf{4.25} & 0.23 & 0.17 & - & - & 0.19 \\
    \rownb & EMT\_Mech~\cite{Sullivan2022} & 136 & 82 & \textbf{DNF} & \textbf{14.01} & 0.27 & 0.20 & - & - & 0.25 \\
    \rownb & macrophage~\cite{DesignPrinciplesGeneNetworks} & 136 & $>1000$ & N/A & 0.54 & 1.09 & 0.84 & - & - & 0.27 \\
    \rownb & angiogenesis~\cite{DesignPrinciplesGeneNetworks} & 141 & $>1000$ & N/A & 0.16 & 1.07 & 1.06 & - & - & 0.16 \\
    \rownb & angiofull~\cite{Weinstein2017} & 142 & $>1000$ & N/A & 0.17 & 1.06 & 0.88 & - & - & 0.23 \\
    \rownb & EMT\_Mech\_TGFbeta~\cite{Sullivan2022} & 150 & 492 & \textbf{DNF} & \textbf{11.28} & 0.78 & 0.69 & - & - & 0.35 \\
    \rownb & RA\_apoptosis~\cite{aghamiri2020automated} & 180 & $>1000$ & N/A & \textbf{DNF} & 1.43 & 1.55 & - & - & 0.19 \\
    \rownb & MAPK~\cite{aghamiri2020automated} & 181 & $>1000$ & N/A & \textbf{13.58} & 1.76 & 1.51 & - & - & 0.27 \\

    \midrule % 200 <= n
    \rownb & Snf1-pathway~\cite{Lubitz2015} & 202 & $>1000$ & N/A & 1.13 & 1.47 & 1.43 & - & - & 0.31 \\
    \rownb & T-cell-co-receptor~\cite{DesignPrinciplesGeneNetworks} & 206 & $>1000$ & N/A & \textbf{DNF} & 1.52 & 2.26 & - & - & 0.35 \\
    \rownb & TcellCheckPoint~\cite{hernandez2020computational} & 218 & $>1000$ & N/A & \textbf{4.99} & NM & 1.96 & - & - & 0.28 \\
    \rownb & Mycobacterium~\cite{DesignPrinciplesGeneNetworks} & 317 & $>1000$ & N/A & 0.42 & 2.36 & \textbf{4.91} & - & - & 0.44 \\
    \rownb & Leishmania~\cite{DesignPrinciplesGeneNetworks} & 342 & $>1000$ & N/A & \textbf{DNF} & 2.56 & \textbf{5.62} & - & - & 0.46 \\
    \rownb & Cholocystokinin~\cite{aghamiri2020automated} & 383 & $>1000$ & N/A & 0.36 & 2.99 & \textbf{4.81} & - & - & 0.37 \\
    \rownb & Alzheimer~\cite{aghamiri2020automated} & 762 & $>1000$ & N/A & \textbf{DNF} & NM & \textbf{18.21} & - & - & 0.79 \\

@@

@@

Although the experimental results show the superiority of our approach to \texttt{mpbn} in general, we however note that there is a model in the \texttt{BBM} repository (with identifier 122) where all the four proposed methods for the new approach did not manage to finish the Petri net conversion before the timeout, whereas \texttt{mpbn} can still handle this model.
The model is not very large but its Boolean functions are rather complicated.
This points to the fact that our current choice of using a BDD-based translation to obtain that Petri net encoding, though it provides a small/efficient ASP might be too costly to handle the complex models.
In such a case, a more \emph{naive} encoding might provide a much larger ASP program, with many redundant rules, but easier/faster to obtain.
The evaluation of the feasibility of such strategy, and of its impact on smaller instances, remains to be done.
Recognizing that a model is locally-monotonic and applying in that specific case dedicated strategies as those of \texttt{mpbn} might also be a partial solution.

It is worth noting that there may be possibly other methods for computing minimal/maximal conflict-free siphons in Petri nets, like the methods for generic siphon computation in the field of Petri nets (see~\cite{DBLP:journals/isci/LiuB16} for a survey about these methods).
Although these approaches do not directly support the minimal/maximal conflict-free siphon computation now, we plan to investigate them in the future. 
They could replace our proposed methods if they give significantly better performance.
However, the current methods appear to already perform very well even on the biggest models we have considered.
